---
title: "Hands-on Exercise 1: A Layered Grammar of Graphics - ggplot2 methods"
author: "Wong Zi Qi"
execute:
  warning: false
  message: false
---

# 1. Getting Started

## 1.1 Installing packages

```{r}
pacman::p_load(tidyverse)
```

## 1.2 Importing data

```{r}
exam_data <- read_csv("../data/Exam_data.csv")
```

# 2. R Graphics VS ggplot

::: panel-tabset
# R Graphics

```{r}
hist(exam_data$MATHS)
```

# ggplot2

```{r}
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_histogram(bins=10, 
                 boundary = 100,
                 color="black", 
                 fill="grey") +
  ggtitle("Distribution of Maths scores")
```
:::

# 3. Essential Grammatical Elements in ggplot2

## 3.1 data

```{r}
ggplot(data=exam_data)
```

::: callout-note
-   A blank canvas appears.
-   `ggplot()` initializes a ggplot object.
:::

## 3.2 Aesthetic Mappings

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS))
```

::: callout-note
-   X-axis is included here.
:::

## 3.3 geom

### geom_bar()

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) + 
  geom_bar()
```

### geom_dotplot()

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(dotsize = 0.5)
```

::: callout-warning
-   Misleading y-scale.
:::

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,         
               dotsize = 0.5) +      
  scale_y_continuous(NULL,           
                     breaks = NULL)  
```

::: callout-note
-   `scale_y_continuous()` is used to turn off the y-axis.
:::

### geom_histogram()

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram()
```

### Modifying geom object

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20,            
                 color="black",      
                 fill="light blue") 
```

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           fill = GENDER)) +
  geom_histogram(bins=20, 
                 color="grey30")
```

### geom_density()

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS, 
           colour = GENDER)) +
  geom_density()
```

### geom_boxplot()

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot(notch=TRUE)
```

::: callout-note
-   Notch is used to help visually assess whether the medians of distributions differ.
:::

### geom_violin()

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_violin()
```

### geom_point()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() 
```

### Combining geom objects

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot() +                    
  geom_point(position="jitter", 
             size = 0.5)  
```

## 3.4 stat

### stat_summary()

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point",       
               fun = "mean",         
               colour ="red",        
               size=4)   
```

### geom_point()

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  geom_point(stat="summary",        
             fun="mean",           
             colour="red",          
             size=4)  
```

### geom_smooth()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5)
```

## 3.5 Facets

### facet_wrap()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_wrap(~ CLASS)
```

### facet_grid()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_grid(~ CLASS)
```

## 3.6 Coordinates

### Flipping axis

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip()
```

### Changing axis range

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
```

## 3.7 Themes

::: panel-tabset
# Gray

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()
```

# Classic

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()
```

# Minimal

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()
```
:::
