---
title: "Hands-on Exercise 4"
author: "Wong Zi Qi"
execute:
  warning: false
  message: false
  freeze: true
---

# *Visualizing Distribution*

## 1. Getting started

```{r}
pacman::p_load(ggdist, ggridges, ggthemes,
               colorspace, tidyverse)
```

```{r}
exam <- read_csv("../data/Exam_data.csv")
```

## 2. Visualising Distribution with Ridgeline Plot

### 2.1 Plotting ridgeline graph: *ggridges*

`geom_density_ridges` first estimates data densities and then draws those using ridgelines.

```{r}
#| code-fold: true


ggplot(exam, 
       aes(x = ENGLISH, 
           y = CLASS)) +
  geom_density_ridges(
    scale = 3,
    rel_min_height = 0.01,
    bandwidth = 3.4,
    fill = lighten("#7097BB", .3),
    color = "white"
  ) +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
    ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```

### 2.2 Varying fill colors along the x axis

`geom_density_ridges_gradient()` allows for varying fill colors. However, they do not allow for alpha transparency in the fill.

```{r}
#| code-fold: true


ggplot(exam, 
       aes(x = ENGLISH, 
           y = CLASS,
           fill = stat(x))) +
  geom_density_ridges_gradient(
    scale = 3,
    rel_min_height = 0.01) +
  scale_fill_viridis_c(name = "Temp. [F]",
                       option = "C") +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
  ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```

### 2.3 Mapping the probabilities directly onto colour

`state_density_ridges()` maps the probabilities directly onto color. The argument `calc_ecdf = TRUE` calculates the empirical cumulative density function.

```{r}
#| code-fold: true


ggplot(exam,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = 0.5 - abs(0.5-stat(ecdf)))) +
  stat_density_ridges(geom = "density_ridges_gradient", 
                      calc_ecdf = TRUE) +
  scale_fill_viridis_c(name = "Tail probability",
                       direction = -1) +
  theme_ridges()
```

### 2.4 Ridgeline plots with quantile lines

`geom_density_ridges_gradient()`colours the ridgeline plot by quantile via `stat(quantile)`.

```{r}
#| code-fold: true

ggplot(exam,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = factor(stat(quantile))
           )) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE, 
    quantiles = 4,
    quantile_lines = TRUE) +
  scale_fill_viridis_d(name = "Quartiles") +
  theme_ridges()
```

Instead of using number to define the quantiles, we can also specify quantiles by cut points such as 2.5% and 97.5% tails to colour the ridgeline plot.

```{r}
#| code-fold: true


ggplot(exam,
       aes(x = ENGLISH, 
           y = CLASS, 
           fill = factor(stat(quantile))
           )) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE, 
    quantiles = c(0.025, 0.975)
    ) +
  scale_fill_manual(
    name = "Probability",
    values = c("#FF0000A0", "#A0A0A0A0", "#0000FFA0"),
    labels = c("(0, 0.025]", "(0.025, 0.975]", "(0.975, 1]")
  ) +
  theme_ridges()
```

## 3. Visualising Distribution with Raincloud Plot

### 3.1 Plotting a Half Eye graph

`stat_halfeye()` plots a Half-Eye graph, which contains a half-density and a slab-interval.

```{r}
#| code-fold: true

ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA)
```

### 3.2 Adding boxplot

Use `geom_boxplot()` to produce a narrow boxplot.

```{r}
#| code-fold: true


ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA)
```

### 3.3 Adding Dot Plots

Add third-geometry layer using `stat_dots()`. This produces a half-dotplot, which is similar to a histogram that indicates the number of samples (number of dots) in each bin. We select side = “left” to indicate we want it on the left-hand side.

```{r}
#| code-fold: true

ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left", 
            justification = 1.2, 
            binwidth = .5,
            dotsize = 2)
```

### 3.4 Finishing touch

Use `coord_flip()` to flip the raincloud chart horizontally to give it the raincloud appearance. Use `theme_economist()` to give the raincloud chart a professional publishing standard look.

```{r}
#| code-fold: true


ggplot(exam, 
       aes(x = RACE, 
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left", 
            justification = 1.2, 
            binwidth = .5,
            dotsize = 1.5) +
  coord_flip() +
  theme_economist()
```

# *Visual Statistical Analysis*

## 1. Visual Statistical Analysis with ggstatsplot

### 1.1 Getting Started

```{r}
pacman::p_load(ggstatsplot, tidyverse)
```

```{r}
exam <- read_csv("../data/Exam_data.csv")

as_tibble(exam)
```

### 1.2 One-sample test: *gghistostats()*

`gghistostats()` is used to to build an visual of one-sample test.

```{r}
#| code-fold: true


set.seed(1234)

gghistostats(
  data = exam,
  x = ENGLISH,
  type = "bayes",
  test.value = 60,
  xlab = "English scores"
)
```

### 1.3 Two-sample mean test: *ggbetweenstats()*

`ggbetweenstats()` is used to build a visual for two-sample mean test.

```{r}
#| code-fold: true


ggbetweenstats(
  data = exam,
  x = GENDER, 
  y = MATHS,
  type = "np",
  messages = FALSE
)
```

### 1.4 Oneway ANOVA Test: *ggbetweenstats()*

`ggbetweenstats()` is used to build a visual for One-way ANOVA test.

-   "ns" → only non-significant
-   "s" → only significant
-   "all" → everything

```{r}
#| code-fold: true


ggbetweenstats(
  data = exam,
  x = RACE, 
  y = ENGLISH,
  type = "p",
  mean.ci = TRUE, 
  pairwise.comparisons = TRUE, 
  pairwise.display = "s",
  p.adjust.method = "fdr",
  messages = FALSE
)
```

### 1.5 Significant Test of Correlation: *ggscatterstats()*

`ggscatterstats()` is used to build a visual for Significant Test of Correlation.

```{r}
#| code-fold: true


ggscatterstats(
  data = exam,
  x = MATHS,
  y = ENGLISH,
  marginal = FALSE,
  )
```

### 1.6 Significant Test of Association (Depedence): *ggbarstats()*

`cut()` is used to bind parameters into n-th class variable. `ggbarstats()` is used to build a visual for Significant Test of Association.

```{r}
#| code-fold: true

exam1 <- exam %>% 
  mutate(MATHS_bins = 
           cut(MATHS, 
               breaks = c(0,60,75,85,100))
)

ggbarstats(exam1, 
           x = MATHS_bins, 
           y = GENDER)
```

## 2. Visualising Models

In this section, it covers visualisation of model diagnostic and model parameters by using parameters package.

### 2.1 Getting Started

```{r}
pacman::p_load(readxl, performance, parameters, see)
```

```{r}
car_resale <- read_xls("../data/ToyotaCorolla.xls", 
                       "data")
car_resale
```

### 2.2 Multiple Regression Model using lm()

Use `lm()` to calibrate a multiple linear regression model.

```{r}
model <- lm(Price ~ Age_08_04 + Mfg_Year + KM + 
              Weight + Guarantee_Period, data = car_resale)

model
```

### 2.3 Model Diagnostic: Checking for multicolinearity

```{r}
check_collinearity(model)
```

```{r}
check_c <- check_collinearity(model)
plot(check_c)
```

### 2.4 Model Diagnostic: Checking normality assumption

```{r}
model1 <- lm(Price ~ Age_08_04 + KM + 
              Weight + Guarantee_Period, data = car_resale)

check_n <- check_normality(model1)

plot(check_n)
```

### 2.5 Model Diagnostic: Checking model for homogeneity of variances

```{r}
check_h <- check_heteroscedasticity(model1)

plot(check_h)
```

### 2.6 Model Diagnostic: Complete check

```{r}
check_model(model1)
```

### 2.7 Visualising Regression Parameters: *see*

`plot()` and `parameters()` are used to visualise the parameters of a regression model.

```{r}
plot(parameters(model1))
```
### 2.8 Visualising Regression Parameters: *ggcoefstats()*

`ggcoefstats()` is used to visualise the parameters of a regression model.

```{r}
ggcoefstats(model1, 
            output = "plot")
```
# *Visualising Uncertainty*

## 1. Getting Started

```{r}
pacman::p_load(plotly, crosstalk, DT, 
               ggdist, ggridges, colorspace,
               gganimate, tidyverse)
```

```{r}
exam <- read_csv("../data/Exam_data.csv")
```

## 2. Visualizing the uncertainty of point estimates: *ggplot2*

Uncertainty, on the other hand, is expressed as standard error, confidence interval, or credible interval.

```{r}
my_sum <- exam %>%
  group_by(RACE) %>%
  summarise(
    n=n(),
    mean=mean(MATHS),
    sd=sd(MATHS)
    ) %>%
  mutate(se=sd/sqrt(n-1))
```

::: callout-note
-   `group_by()` of dplyr package is used to group the observation.
-   `summarise()` is used to compute the count of observations, mean, standard deviation.
-   `mutate()` is used to derive standard error.
-   the output is saved as a tibble data table.
:::

```{r}
knitr::kable(head(my_sum), format = 'html')
```
### 2.1 Plotting standard error bars of point estimates

```{r}
#| code-fold: true


ggplot(my_sum) +
  geom_errorbar(
    aes(x=RACE, 
        ymin=mean-se, 
        ymax=mean+se), 
    width=0.2, 
    colour="black", 
    alpha=0.9, 
    linewidth=0.5) +
  geom_point(aes
           (x=RACE, 
            y=mean), 
           stat="identity", 
           color="red",
           size = 1.5,
           alpha=1) +
  ggtitle("Standard error of mean maths score by rac")
```
### 2.2 Plotting confidence interval of point estimates

```{r}
#| code-fold: true


ggplot(my_sum) +
  geom_errorbar(
    aes(x=reorder(RACE, -mean), 
        ymin=mean-1.96*se, 
        ymax=mean+1.96*se), 
    width=0.2, 
    colour="black", 
    alpha=0.9, 
    linewidth=0.5) +
  geom_point(aes
           (x=RACE, 
            y=mean), 
           stat="identity", 
           color="red",
           size = 1.5,
           alpha=1) +
  labs(x = "Maths score",
       title = "95% confidence interval of mean maths score by race")
```
### 2.3 Visualizing the uncertainty of point estimates with interactive error bars

```{r}
#| code-fold: true


shared_df = SharedData$new(my_sum)

bscols(widths = c(4,8),
       ggplotly((ggplot(shared_df) +
                   geom_errorbar(aes(
                     x=reorder(RACE, -mean),
                     ymin=mean-2.58*se, 
                     ymax=mean+2.58*se), 
                     width=0.2, 
                     colour="black", 
                     alpha=0.9, 
                     size=0.5) +
                   geom_point(aes(
                     x=RACE, 
                     y=mean, 
                     text = paste("Race:", `RACE`, 
                                  "<br>N:", `n`,
                                  "<br>Avg. Scores:", round(mean, digits = 2),
                                  "<br>95% CI:[", 
                                  round((mean-2.58*se), digits = 2), ",",
                                  round((mean+2.58*se), digits = 2),"]")),
                     stat="identity", 
                     color="red", 
                     size = 1.5, 
                     alpha=1) + 
                   xlab("Race") + 
                   ylab("Average Scores") + 
                   theme_minimal() + 
                   theme(axis.text.x = element_text(
                     angle = 45, vjust = 0.5, hjust=1)) +
                   ggtitle("99% Confidence interval of average /<br>maths scores by race")), 
                tooltip = "text"), 
       DT::datatable(shared_df, 
                     rownames = FALSE, 
                     class="compact", 
                     width="100%", 
                     options = list(pageLength = 10,
                                    scrollX=T), 
                     colnames = c("No. of pupils", 
                                  "Avg Scores",
                                  "Std Dev",
                                  "Std Error")) %>%
         formatRound(columns=c('mean', 'sd', 'se'),
                     digits=2))
```

## 3. Visualising Uncertainty: *ggdist*

`ggdist` is designed for both frequentist and Bayesian uncertainty visualization, taking the view that uncertainty visualization can be unified through the perspective of distribution visualization:

-   for frequentist models, one visualises confidence distributions or bootstrap distributions (see vignette(“freq-uncertainty-vis”));
-   for Bayesian models, one visualises probability distributions (see the tidybayes package, which builds on top of ggdist).

### 3.1 Visualizing the uncertainty of point estimates: *ggdist*

`stat_pointinterval()` is used to build a visual for displaying distribution.

```{r}
#| code-fold: true


exam %>%
  ggplot(aes(x = RACE, 
             y = MATHS)) +
  stat_pointinterval() +
  labs(
    title = "Visualising confidence intervals of mean math score",
    subtitle = "Mean Point + Multiple-interval plot")
```
Below are some arguments that can be adjusted:

-   .width
-   .point
-   .interval

```{r}
#| code-fold: true


exam %>%
  ggplot(aes(x = RACE, y = MATHS)) +
  stat_pointinterval(.width = 0.95,
  .point = median,
  .interval = qi) +
  labs(
    title = "Visualising confidence intervals of median math score",
    subtitle = "Median Point + Multiple-interval plot")
```
The plot is modified to 95% and 99% CI.

```{r}
#| code-fold: true

exam %>%
  ggplot(aes(x = RACE, 
             y = MATHS)) +
  stat_pointinterval(.width = c(0.95, 0.99),
    show.legend = FALSE) +   
  labs(
    title = "Visualising confidence intervals of mean math score",
    subtitle = "Mean Point + Multiple-interval plot")
```

### 3.2 Varying fill colors

`stat_gradientinterval()` is used to build a visual for displaying distribution.

```{r}
#| code-fold: true


exam %>%
  ggplot(aes(x = RACE, 
             y = MATHS)) +
  stat_gradientinterval(.width = c(0.95, 0.99),   
    fill = "skyblue",      
    show.legend = TRUE     
  ) +                        
  labs(
    title = "Visualising confidence intervals of mean math score",
    subtitle = "Gradient + interval plot")
```

## 4. Visualising Uncertainty with Hypothetical Outcome Plots (HOPs)

### 4.1 Getting Started

```{r}
devtools::install_github("wilkelab/ungeviz")
```

```{r}
library(ungeviz)
```

### 4.2 Plotting HOPs

```{r}
#| code-fold: true


ggplot(data = exam, 
       (aes(x = factor(RACE), 
            y = MATHS))) +
  geom_point(position = position_jitter(
    height = 0.3, 
    width = 0.05), 
    size = 0.4, 
    color = "#0072B2", 
    alpha = 1/2) +
  geom_hpline(data = sampler(25, 
                             group = RACE), 
              height = 0.6, 
              color = "#D55E00") +
  theme_bw() + 
  transition_states(.draw, 1, 3)
```
# *Funnel Plots for Fair Comparisons*

## 1. Getting Started

```{r}
pacman::p_load(tidyverse, FunnelPlotR, plotly, knitr)
```

```{r}
covid19 <- read_csv("../data/COVID-19_DKI_Jakarta.csv") %>%
  mutate_if(is.character, as.factor)

covid19
```

## 2. FunnelPlotR methods

It requires a `numerator` (events of interest), `denominator` (population to be considered) and `group`. The key arguments selected for customisation are:

-   `limit`: plot limits (95 or 99).
-   `label_outliers`: to label outliers (true or false).
-   `Poisson_limits`: to add Poisson limits to the plot.
-   `OD_adjust`: to add overdispersed limits to the plot.
-   `xrange` and `yrange`: to specify the range to display for axes, acts like a zoom function.
-   Other aesthetic components such as graph title, axis labels etc.

### 2.1 The basic plot

```{r}
#| code-fold: true


funnel_plot(
  .data = covid19,
  numerator = Positive,
  denominator = Death,
  group = `Sub-district`
)
```

::: callout-note
-   `group` in this function is different from the scatterplot. Here, it defines the level of the points to be plotted i.e. Sub-district, District or City. If Cityc is chosen, there are only six data points.
-   By default, `data_type` argument is “SR”.
-   `limit`: Plot limits, accepted values are: 95 or 99, corresponding to 95% or 99.8% quantiles of the distribution.
:::

### 2.2 Makeover 1

```{r}
#| code-fold: true


funnel_plot(
  .data = covid19,
  numerator = Death,
  denominator = Positive,
  group = `Sub-district`,
  data_type = "PR",     #<<
  xrange = c(0, 6500),  #<<
  yrange = c(0, 0.05)   #<<
)
```

::: callout-note
-   `data_type` argument is used to change from default “SR” to “PR” (i.e. proportions).
-   `xrange` and `yrange` are used to set the range of x-axis and y-axis.
:::

### 2.3 Makeover 2

```{r}
#| code-fold: true


funnel_plot(
  .data = covid19,
  numerator = Death,
  denominator = Positive,
  group = `Sub-district`,
  data_type = "PR",   
  xrange = c(0, 6500),  
  yrange = c(0, 0.05),
  label = NA,
  title = "Cumulative COVID-19 Fatality Rate by Cumulative Total Number of COVID-19 Positive Cases", #<<           
  x_label = "Cumulative COVID-19 Positive Cases", #<<
  y_label = "Cumulative Fatality Rate"  #<<
)
```


## 3. Funnel Plot for Fair Visual Comparison: *ggplot2*

### 3.1 Computing the basic derived fields

To plot the funnel plot from scratch, we need to derive cumulative death rate and standard error of cumulative death rate.

```{r}
df <- covid19 %>%
  mutate(rate = Death / Positive) %>%
  mutate(rate.se = sqrt((rate*(1-rate)) / (Positive))) %>%
  filter(rate > 0)
```

```{r}
fit.mean <- weighted.mean(df$rate, 1/df$rate.se^2)
```

### 3.2 Calculate lower and upper limits for 95% and 99.9% CI

```{r}
number.seq <- seq(1, max(df$Positive), 1)
number.ll95 <- fit.mean - 1.96 * sqrt((fit.mean*(1-fit.mean)) / (number.seq)) 
number.ul95 <- fit.mean + 1.96 * sqrt((fit.mean*(1-fit.mean)) / (number.seq)) 
number.ll999 <- fit.mean - 3.29 * sqrt((fit.mean*(1-fit.mean)) / (number.seq)) 
number.ul999 <- fit.mean + 3.29 * sqrt((fit.mean*(1-fit.mean)) / (number.seq)) 
dfCI <- data.frame(number.ll95, number.ul95, number.ll999, 
                   number.ul999, number.seq, fit.mean)
```

### 3.3 Plotting a static funnel plot

```{r}
#| code-fold: true


p <- ggplot(df, aes(x = Positive, y = rate)) +
  geom_point(aes(label=`Sub-district`), 
             alpha=0.4) +
  geom_line(data = dfCI, 
            aes(x = number.seq, 
                y = number.ll95), 
            size = 0.4, 
            colour = "grey40", 
            linetype = "dashed") +
  geom_line(data = dfCI, 
            aes(x = number.seq, 
                y = number.ul95), 
            size = 0.4, 
            colour = "grey40", 
            linetype = "dashed") +
  geom_line(data = dfCI, 
            aes(x = number.seq, 
                y = number.ll999), 
            size = 0.4, 
            colour = "grey40") +
  geom_line(data = dfCI, 
            aes(x = number.seq, 
                y = number.ul999), 
            size = 0.4, 
            colour = "grey40") +
  geom_hline(data = dfCI, 
             aes(yintercept = fit.mean), 
             size = 0.4, 
             colour = "grey40") +
  coord_cartesian(ylim=c(0,0.05)) +
  annotate("text", x = 1, y = -0.13, label = "95%", size = 3, colour = "grey40") + 
  annotate("text", x = 4.5, y = -0.18, label = "99%", size = 3, colour = "grey40") + 
  ggtitle("Cumulative Fatality Rate by Cumulative Number of COVID-19 Cases") +
  xlab("Cumulative Number of COVID-19 Cases") + 
  ylab("Cumulative Fatality Rate") +
  theme_light() +
  theme(plot.title = element_text(size=12),
        legend.position = c(0.91,0.85), 
        legend.title = element_text(size=7),
        legend.text = element_text(size=7),
        legend.background = element_rect(colour = "grey60", linetype = "dotted"),
        legend.key.height = unit(0.3, "cm"))
p
```

### 3.4 Interactive Funnel Plot: *plotly* + *ggplot2*

```{r}
#| code-fold: true


fp_ggplotly <- ggplotly(p,
  tooltip = c("label", 
              "x", 
              "y"))
fp_ggplotly
```



